---
title: "MSc Bioinformatics Coursework 2024/25"
author: "Kehinde Oyesola"
date: "02/11/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#set up the environment, load packages and data needed for coursework
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(MASS)
options(bitmapType='cairo')
rm(list=ls())
load("assess_data_1224.Rdata")
```

# 1 Simulate Markov Chains

## Q1 Comment Code

```{r}
#Defining parameters

#set seed to allow any randomly generated data to be reproducible
set.seed(42)

# the probability of going from state 1 to state 1
a <- 0.8999 

# the probability of going from state 2 to state 2
b <- 0.69 

#number of days to simulate
n_days <- 700 

#creates a 3x3 matrix of the transition probabilties 
transition_matrix <- matrix(c(a, 0.1, 0.0001,
                              0.3, b, 0.01,
                              0, 0, 1), nrow=3, ncol=3, byrow=TRUE)

#the initial state to start from i.e state 1
state <- 1 

#vector to store the simulated values over 700days
patient_record <- rep(0, n_days) 

#for loop to iterate 700 times
for (day in 1:n_days) { 
   # gets the probabilities in the row of the matrix according to the state
  pr <- transition_matrix[state, ]
  
  #sample from numbers 1-3 based on the probabilities and update the state with the value chosen
  state <- sample(c(1:3), size = 1, prob = pr) 
  
  #for each iteration, store the sampled state into the corresponding index position in the vector
  patient_record[day] <- state 
}


#display the simulated patient record over the 700 days on a line plot
plot(1:n_days, patient_record, "l") 
```

## Q2 Extend the Markov Chain

```{r}
#Defining parameters
set.seed(42) 
a <- 0.8999 
b <- 0.29 

# the probability of going from state 3 to state 3
c <- 0.55 

n_days <- 700 

#creates a 4x4 matrix of the transition probabilties 
transition_matrix <- matrix(c(a, 0.1, 0, 0.0001,
                              0.3, b, 0.4, 0.01,
                              0.2, 0.2, c, 0.05, 0, 0, 0, 1), 
                            nrow=4, ncol=4, byrow=TRUE)

state <- 1  

patient_record <- rep(0, n_days)


for (day in 1:n_days) { 
  pr <- transition_matrix[state, ] 
  
  #sample from numbers 1-4 based on the probabilities and update the state
  state <- sample(c(1:4), size = 1, prob = pr) 
  
  patient_record[day] <- state 
}

plot(1:n_days, patient_record, "l") 
```

### Analysis of Q2 results

There is no transition from state 4 (dead) to the other states except to state 4 because the probability of transitioning from state 4 to other states is 0, whereas transition from state 4 to state 4 has a probability of 1. Therefore, from day 23 where state 4 was first sampled, it will remain so indefinitely for the rest of the simulated days.


## Q3 Repeat the Simulation

Repeat the simulation for the four-state Markov Chain above for 1000 patients, by extending your code from the previous question or otherwise. Choose an appropriate plot to display how long a patient spends in each state, on average. Also, calculate the average time spent in each state for the 1000 patients. Explain the results and what the selected plot shows.

### Repeat the simulation for the four-state Markov Chain above for 1000 patients, by extending your code from the previous question or otherwise.
```{r}
#Defining parameters
set.seed(42) #set seed to allow any randomly generated data to be reproducible
a <- 0.8999 # the probability of going from state 1 to state 1
b <- 0.29 # the probability of going from state 2 to state 2
c <- 0.55 # the probability of going from state 3 to state 3

n_days <- 700 #number of days to simulate

n_patients <- 1000 #number of patients

#creates a 4x4 matrix of the transition probabilities 
transition_matrix <- matrix(c(a, 0.1, 0, 0.0001,
                              0.3, b, 0.4, 0.01,
                              0.2, 0.2, c, 0.05, 0, 0, 0, 1), 
                            nrow=4, ncol=4, byrow=TRUE)

data <- matrix(0, nrow=n_patients, ncol=n_days) #matrix to store simulated states for each patient

#vector to store row names of the matrix
matrixrow<- rep(0, n_patients)

#vector to store column names of the matrix
matrixcol <- rep(0, n_days) 

#change each element in the vector to Patient 1, Patient 2 etc
for (i in 1: n_patients){
  matrixrow[i] <- paste("Patient", i)
}

#change each element in the vector to Day 1, Day 2 etc
for (i in 1: n_days){
  matrixcol[i] <- paste("Day", i)
}

#change the data matrix row and column names
rownames(data) <- matrixrow
colnames(data) <- matrixcol

#for loop to iterate for 1000 patients
for (p in 1:n_patients){
  state <- 1 
  
  # nested for loop to iterate over 700 days
  for (day in 1:n_days) { 
    pr <- transition_matrix[state, ] 
    
    state <- sample(c(1:4), size = 1, prob = pr) 
    
    #for each iteration, store the sampled state into the corresponding index position in the matrix
    data[p,day] <- state 
  }
  
}


```

### Choose an appropriate plot to display how long a patient spends in each state, on average. 

Code to calculate how long a patient spends in each state
```{r}
states <- c(1:4) #initialise states

#initialise empty data frame to store no of consecutive days each patient spends in each state
days_in_state <- data.frame(Patient_no = integer(), 
                            State = integer(), 
                            Duration = integer())


for (record in 1:n_patients) { #iterate over each patient

    # extract individual pateint record in the matrix
    patient_record <- data[record, ]
    
    # count consecutive days the patient is in each state using rle() function
    con_days <- rle(patient_record) 
    
    
    for (s in states){ #iterate over the different states
      
      if (s %in% con_days$values) { 
        
      # Find the values of consecutive days for each state if the state is present
      con_days_per_state <- con_days$lengths[con_days$values == s]
      
      # Add each duration to the dataframe
      for (days in con_days_per_state) {
        days_in_state <- rbind(days_in_state,
                                 data.frame(Patient_no = record, State = s, Duration = days))
      }
    } else {
      
      # If the state is not present, record a duration of 0
      days_in_state <- rbind(days_in_state,
                               data.frame(Patient_no = record, State = s, Duration = 0))
    }
  }
}    


#Empty matrix to store how long a patient spends in each state on average
average_con_days <- matrix(0, nrow=1, ncol= 4)
  
for (s in states) {
  #filter the dataframe for each patient and each state and duration
      total <- days_in_state[days_in_state$State == s, "Duration"] 
      
      # find the mean of the filtered columns and round to 2 decimal places
      average_con_days[s] <- round(mean(total)) 
      
}

```

Plot to display how long a patient spends in each state, on average. 

```{r}
#convert average_con_days to a dataframe
average_con_days_df <- as.data.frame(average_con_days)
colnames(average_con_days) <- c(1:4)
#group the values according to states
average_con_days_df <- pivot_longer(
  data = average_con_days_df,
  cols = everything(),    # Convert all columns to key-value pairs
  names_to = "State",     # Column for state names
  values_to = "Average"   # Column for average consecutive days
)

ggplot(average_con_days_df, aes(x = State, y = Average, fill = State)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Time spent in each state by a patient on average",
    x = "State",
    y = "Days",
    fill = "State"
  ) + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) + scale_color_brewer(palette="Set1")
```


Also, calculate the average time spent in each state for the 1000 patients. 
```{r}
#Empty matrix to store how long a patient spends in each state on average
average_per_patient <- matrix(0, nrow= n_patients, ncol= 4)


for (s in states) {
    for (p in 1:n_patients){
      total <- days_in_state[days_in_state$Patient_no == p & days_in_state$State == s, "Duration"] #filter the dataframe for each patient and each state and duration
      
      average_per_patient[p,s] <- round(mean(total)) # find the mean of the filtered columns and round to 2 decimal places
  }   
}

#convert average_per_patient to a dataframe
average_per_patient_df <- as.data.frame(average_per_patient)

#change column names to states
colnames(average_per_patient_df) <- c(1:4)

#create new row with patient numbers variable already declared in previous chunk
average_per_patient_df$Patient_no <- c(1:100)


#group the values according to states
average_per_patient_df <- pivot_longer(
  data = average_per_patient_df,
  cols = -Patient_no,    # Convert all columns to key-value pairs except patient_no column
  names_to = "State",     # Column for state names
  values_to = "Average"   # Column for average consecutive days
)


ggplot(average_per_patient_df, aes(x = Patient_no, y = Average, fill = State)) +   #geom_bar(stat = "identity", position = "stack") +
  geom_point() + geom_jitter()+
  labs(
    title = "Average time spent in each state for the 1000 patients",
    x = "Patient",
    y = "Days",
    fill = "State"
  ) + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) #+ scale_color_brewer(palette="Set1")

#boxplot
ggplot(average_per_patient_df, aes(x = Patient_no, y = Average, fill = State)) +   #geom_bar(stat = "identity", position = "stack") +
  geom_point() 

ggplot(average_per_patient_df, aes(State, y=Average)) + geom_boxplot()
```

Explain the results and what the selected plot shows.

On average patients spend more time in state 1 and 4 than in state 3.

# 2 Data Analysis

## Q 5 Study the following code example and add comments to describe what it does

```{r}

#load necessary packages
library(tidyverse) 
library(MASS)

#load the dataset required
load("data/assess_data_1224.Rdata")

#assign new variables
idx <- 20 
c_cl <- 15:30

#plot a boxplot of log2 of the Y matrix of all rows from column 15 to 30. 
#Plus 1 as you can't log a 0.
boxplot(log2(Y[, c_cl] + 1)) 

#extract data in row 15 to 30 in tissue column in patient_data  
x <- patient_data$tissue[c_cl] 

#extract data in row 15 to 30 in patient column in patient_data 
z <- patient_data$patient[c_cl] 

#create a dataframe using specific values
tmp <- data.frame(y = Y[idx, c_cl], x = x, z = z, lib_size = colSums(Y[, c_cl])) 

#perform general linear model based on poisson distribution rule
out <- glm(y ~ x + z + lib_size, data = tmp, family = "poisson") 

#extract the only the coefficients from the summary
p_val <- summary(out)$coefficients[2, 4] 

```

# 2.1 Data Exploration

## Q 6 Quality Control

There is one problematic sample in the data, identify the problematic sample and explain why it is problematic. Provide evidence and/or a plot to support your answer. You could consider using some statistics on the data to identify the problematic sample or any other suitable approach.

```{r}

#get summary of the Y matrix
summary(Y)


boxplot(log2(Y+ 1))

```
Looking at the summary for each sample and the boxplots, Sample 3 is the problematic sample because it has a gene expression level of zero for all genes. 


## Q 7 Differential Expression

Using the code from Q5, perform a regression-based differential expression analysis between all normal and tumour samples using Poisson regression. Use the tissue type as the only covariate. Plot the appropriate log10 p-value from your analysis.

```{r}

#assign data tissue column in patient_data to x
x <- patient_data$tissue  

#get the no of genes in the matrix
g_rows <- nrow(Y) 

#initialise empty vector to store pvalues of each gene 
p_val <- rep(0, g_rows) 

# Loop over rows(genes) in y matrix and perform glm
for (gene in 1:g_rows) {
  
    #create a dataframe of each row of genes in Y and x
    tmp1 <- data.frame(y = Y[gene, ], x = x) 
    
    #perform general linear model based on poisson distribution rule
    out1 <- glm(y ~ x, data = tmp1, family = "poisson") 
    
    #extract pvalues for each gene
    p_val[gene] <- summary(out1)$coefficients[2, 4]
}

#adjust the p values using Benjamini-Hochberg method
adj_p_val <- p.adjust(p_val, "BH")

# create data.frame of the adjusted p-values and gene names 
adj_p_val_df <- data.frame(p_values = adj_p_val, gene_index = 1:g_rows) 

#plot each pvalue for each gene
ggplot(adj_p_val_df, aes(x = gene_index, y = -log10(p_values+1))) +
    geom_point(size = 1, col = "red") +
    labs(x = "Gene Index" ,
         y = "-log10(adjusted p-value)",
         title = "Manhattan Plot of Gene Expression") +
    theme(legend.title = element_blank(), plot.title=element_text(hjust=0.5))

```

##Q 8 Differential Expression with Covariates

Perform a regression-based analysis to identify genes differentially expressed between normal and tumour samples including the tissue variable indicating if it is tumour or normal sample. Plot the appropriate log10 p-value from your analysis.

Compare the p-values with and without inclusion of the tissue type as a covariate, what do you observe? Which of the covariate has the biggest effect? Explain your answer with supporting plots, tables and further analysis if required. 

```{r}
# Tissue variable
x <- patient_data$tissue  

# Patient ID variable
z <- patient_data$patient_id  

# Number of genes
g <- nrow(Y)  

# Create empty vectors to store results
gene_name <- rep(0, g)


#empty vector to store pvalues from glm without tissue covariate
pval1 <- matrix(0, nrow = g, ncol = no_of_coef1) 

# empty vector to store pvalues from glm without tissue covariate
pval2 <- matrix(0, nrow = g, ncol = no_of_coef2)

# empty vector to store estimates from glm with tissue covariate
estimate1 <- matrix(0, nrow = g, ncol = no_of_coef1) 

# empty vector to store estimates from glm without tissue covariate
estimate2 <- matrix(0, nrow = g, ncol = no_of_coef2) 

#empty vector to store chi square pvalues from annova test
chisq_pval <- rep(0, g)

#empty vector to store deviance values from annova test
chisq_deviance <- rep(0, g)

# Loop over genes 
for (gene in 1:g) {
  
  #give updates for every 100 genes processed
  if (gene %% 100 == 0) cat("Processing gene:", gene, "\n")
  
  #create data frame 
  # Exclude column 3 and 18 from Y and x, z since 3 is problematic
    tmp3 <- data.frame(y = Y[gene, c(-3,-18)], x = x[c(-3,-18)], z = z[c(-3,-18)], lib_size = colSums(Y[, c(-3,-18)]))  
    
    # perform glm with and without tissue type as a covariate 
    out3 <- list(
        glm(y ~ x + z + lib_size, data = tmp3, family = "poisson"),
        glm(y ~ z + lib_size, data = tmp3, family = "poisson")
    )
    
    # Get number of coefficients for glm
    coef_no1 <- nrow(summary(out3[[1]])$coefficients)
    coef_no2 <- nrow(summary(out3[[2]])$coefficients)
    
    # Extract p-values and estimates for each glm and store in matrices
    for (coef in 2:coef_no1) {
        pval1[gene, coef] <- summary(out3[[1]])$coefficients[coef, 4]  
        estimate1[gene, coef] <- summary(out3[[1]])$coefficients[coef, 1]
    }
    
    for (coef in 2:coef_no2) {
        pval2[gene, coef] <- summary(out3[[2]])$coefficients[coef, 4]  
        estimate2[gene, coef] <- summary(out3[[2]])$coefficients[coef, 1]
    }
    
    # Store gene name in gene_name matrix
    gene_name[gene] <- rownames(Y)[gene]  
    
    # Perform Annova Likelihood Ratio Test (LRT) on both glm
    lrt_test <- anova(out3[[1]], out3[[2]], test = "LRT")
    
    # Extract pvalue and deviance from the annova test
    chisq_pval[gene] <- lrt_test[2, 5]  
    chisq_deviance[gene] <- lrt_test[2, 4]
}

load("covariates2.RData")

#change column names according to coefficients in glm
coeff1_names<- rownames(summary(out3[[1]])$coefficients)
colnames(estimate1) <- coeff1_names
colnames(pval1) <- coeff1_names
coeff2_names<- rownames(summary(out3[[2]])$coefficients)
colnames(estimate2) <- coeff2_names
colnames(pval2) <- coeff2_names


#new no of coeff to exclude the intercept column
coef_no1_1 <- coef_no1-1
coef_no2_1 <- coef_no2-1

#empty matrices to store adjusted pvlaues of both glm
adj_pvals1 <- matrix(0, nrow = g , ncol = coef_no1_1)
adj_pvals2  <- matrix(0, nrow = g , ncol = coef_no2_1)

#adjust p-values using the Benjamini-Hochberg method for both glm
for (val in 2:coef_no1){
  adj_pvals1[,val-1] <- p.adjust(pval1[,val], "BH")
                          
}

for (val in 2:coef_no2){
  adj_pvals2[,val-1] <- p.adjust(pval2[,val], "BH")
                          
}


#data frames to store adjusted pvalues
pval1_df <- data.frame(p_vals = adj_pvals1, gene_index = 1:g, gene_names = gene_name)
pval2_df <- data.frame(p_vals = adj_pvals2, gene_index = 1:g, gene_names = gene_name)

#data frames to store the estimates
estimates1_df <- data.frame(gene_index = 1:g, estimates = estimate1, gene_names = gene_name)
estimates2_df <- data.frame(gene_index = 1:g, estimates = estimate2, gene_names = gene_name)



#save.image(file = "covariates.RData")
```



Compare the p-values with and without inclusion of the tissue type as a covariate, what do you observe? Which of the covariate has the biggest effect? Explain your answer with supporting plots, tables and further analysis if required. 

```{r}


#empty matrix to store average pvalues and estimates
average_est <- matrix(0, nrow = coef_no2 , ncol = 2)
average_pval <- matrix(0, nrow = coef_no2 , ncol = 2)

#calculate the mean estimates and pvalues for each coefficicent in both glm
#excludes first column which is the intercept
for (val in 2:coef_no1){
  average_est[val-1,1] <- mean(estimate1[, val])
  average_pval[val-1,1] <- mean(pval1_df[, val])
                          
}

for (val in 2:coef_no2){
  average_est[val,2] <- mean(estimate2[, val])
  average_pval[val,2] <- mean(pval2_df[, val])
}



#dataframe to store the average estimates and pvalues and specifcy if tissue type(x) is included in the glm
ave1_df <- data.frame(coefficients = coeff1_names[-1], 
                          estimates = average_est[1:coef_no1_1,1], 
                          pvals = average_pval[1:coef_no1_1,1], 
                          with_x = rep("Yes", coef_no1_1))

ave2_df <- data.frame(coefficients = coeff2_names[-1], 
                          estimates = average_est[2:coef_no1_1,2], 
                          pvals = average_pval[2:coef_no1_1,2], 
                          with_x = rep("No", coef_no2_1))

#combine average estimates and pvalues from both glm
combined_df <- bind_rows(ave1_df, ave2_df)

#plot bar chart to compare estimates from both glm
ggplot(combined_df, aes(y = estimates, x = coefficients, fill = with_x))+
  geom_bar(stat = "identity", position = "dodge") + ylim(-1, 1)+
  labs(title = "Comparison of Estimates",
    x = "Coefficients",
    y = "Estimates",
    fill = "Tissue type included in Glm?") +
  theme(plot.title=element_text(hjust=0.5, size = 10), legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 0.5))

#plot bar chart to compare -log10(pvals) from both glm
ggplot(combined_df, aes(y = -log10(pvals), x = coefficients, fill = with_x))+
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Comparison of Pvalues",
    x = "Coefficients",
    y = "-log10(pvalues)",
    fill = "Tissue type included in Glm?") +
  theme(plot.title=element_text(hjust=0.5), legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 0.5))

boxplot(-log10(pvals) ~ with_x, combined_df)
boxplot(estimates ~ with_x, combined_df)
mean(ave1_df$estimates)

mean(ave2_df$estimates)
```

The pvalues of patient_id covariates and the lib_size, with the inclusion of tissue type as a covariate is not different to the pvalues without the inclusion of tissue type as a covariate. 

However, when tissue type is included as a covariate, the average effect size of the patient_id increases on average. 

The tissue type has the biggest effect, followed by the patient id. The library size had very minimal effect with coefficient values of 2.923875e-08 and 2.649152e-08 with and without tissue type respectively as a covariate. Interestingly, the pvalues of the lib_size is extremely significant, however due to the extremely low effect size, the lib_size is directly related to the 


```{r}
ggplot(combined_df, aes(x = log2(estimates), y = -log10(p_vals))) +
  geom_point(aes(color = Significance), alpha = 0.8, size = 1) +
  scale_color_manual(values = c("green", "red")) +
  theme_minimal() +
  labs(title = "Volcano Plot",
       x = "Log2 Fold Change",
       y = "-Log10 Adjusted P-value") +
  theme(legend.title = element_blank(), plot.title=element_text(hjust=0.5), axis.text.x = element_text(angle = 90, hjust = 0.5)) +
  geom_vline(xintercept = c(-1, 1), col = "black", linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), col = "black", linetype = "dashed") +
  geom_text(data = p_val_df[1:10, ],  # Label top 10 significant genes
            aes(label = gene_no), size = 3, vjust = 1.5, hjust = 0.5)
```




```{r}

#data frame to store chi-squared p-values from annova LRT
chsq_df <- data.frame(gene_index = 1:g, chisq_pvals = chisq_pval, deviance = chisq_deviance,  gene_names = gene_name)

#view summary of chi square p values and deviance
summary(chsq_df$chisq_pvals)
summary(chsq_df$deviance)

#plot the -log10 pvalues +1 
ggplot(chsq_df, aes(y = -log10(chisq_pvals+1), x = gene_index))+
  geom_point(colour = "red")+
  labs(title = "Chi squared pValues to show significance of the tissue variable", 
       x= "gene index",
       y= " -log10 chi squared pvalues")+ 
   theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

```

The average chi squared pvalues when the both model was compared is 0.005.
The average defiance for when tissue variable is excluded is -69473. This suggests excluding the tissue type variable create a worse fit than when it is included. Suggesting that the tissue type variable is statistically significant.